---
title: "Building a muliple linear regression model for predicting flight landing distance"
author: "Puneet Bhatia"
date: "1/21/2020"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---



```{r results='hide',message=FALSE,warning=FALSE}
#Loading required libraries

library(readxl)
library(tidyverse)
library(plyr)
library(dplyr)
library(MASS)

```

__Initial exploration of the data__

```{r}


# Importing data

FAA1 <- read_excel("FAA1-1.xls") 
FAA2 <- read_excel("FAA2-1.xls") 

# Structure of datasets

str(FAA1)

dim(unique(FAA1))

```

FAA1 has 800 rows and 8 columns

```{r}
str(FAA2)
dim(unique(FAA2))
```

FAA2 has 150 rows and 7 columns. 'duration' column is not present in FAA2, but it is present in FAA1.

```{r}
# Appending the data

combined <- rbind.fill(FAA1,FAA2)

# Checking duplicates

dim(combined[,-2])

dim(unique(combined[,-2]))

```

We have 100 duplicate records.

```{r}
# Removing duplicate records

unique_combined <- unique(combined[,-2]) %>% left_join(FAA1)

# Structure of the new data set

str(unique_combined)

```

Dataset has 850 rows and 8 variables

```{r}

#unique_combined$aircraft <- as.factor(unique_combined$aircraft)
# Summary of the combined data

summary(unique_combined)

boxplot(unique_combined$duration~unique_combined$aircraft)
boxplot(unique_combined$speed_ground~unique_combined$aircraft)
boxplot(unique_combined$speed_air~unique_combined$aircraft)
boxplot(unique_combined$height~unique_combined$aircraft)
boxplot(unique_combined$pitch~unique_combined$aircraft)
boxplot(unique_combined$distance~unique_combined$aircraft)
boxplot(unique_combined$no_pasg~unique_combined$aircraft)

hist(unique_combined$distance)
```

__Summary of findings__

* There are 642 (~75.5%) missing values for spped_air and 50 (~5.88%) missing values for duration. Other columns do not have missing values 
* Distance has a right skewed distribution. Median is much smaller than the mean and this can also be seen from the histogram
* Boeing has relatively higher distribution of distance and pitch than airbus
* Boeing has more speed_air variation. Mean and median are very close for both aircrafts
* Duration, No_pasg, Speed_ground and Height have similar distribution for both the aircrafts

__Data Cleaning and further exploration__

```{r}
# Checking abnormal values

unique_combined %>% filter(duration<40)  #5 observations with duration less than 40
unique_combined %>% filter(!(speed_ground >= 30 & speed_ground <= 140)) # 3 observations with abnormal speed_ground
unique_combined %>% filter(!(speed_air >= 30 & speed_air <= 140| is.na(speed_air))) # 1 observation with abnormal speed_air
unique_combined %>% filter(height<6)  #10 observations with abnormal height
unique_combined %>% filter(distance>6000)  #2 observations with abnormal distance
```

```{r}
#Cleaning data based on requirements

FAA_final <- unique_combined %>% filter(duration>40 | is.na(duration),speed_ground >= 30 & speed_ground <= 140, speed_air >= 30 & speed_air <= 140 | is.na(speed_air), height >= 6, distance < 6000)

dim(FAA_final)
```

The dataset has 831 rows after removing the abnormal values. 19 rows have been removed.

```{r}
# Structure of cleaned data

str(FAA_final)

summary(FAA_final)

# Histogram of the variables to visualize the distribution

hist(FAA_final$no_pasg)
hist(FAA_final$speed_ground)
hist(FAA_final$speed_air)
hist(FAA_final$height)
hist(FAA_final$pitch)
hist(FAA_final$distance)
hist(FAA_final$duration)

# Boxplots of the numerical variables

boxplot(FAA_final$duration~FAA_final$aircraft)
boxplot(FAA_final$speed_ground~FAA_final$aircraft)
boxplot(FAA_final$speed_air~FAA_final$aircraft)
boxplot(FAA_final$height~FAA_final$aircraft)
boxplot(FAA_final$pitch~FAA_final$aircraft)
boxplot(FAA_final$distance~FAA_final$aircraft)
boxplot(FAA_final$no_pasg~FAA_final$aircraft)


```

__Summary of findings__

* There are 628 (~75.5%) missing values for speed_air and 50 (~6.01%) missing values for duration. Other columns do not have missing values 
* no_pasg, speed_ground, hight and pitch have symetric distribution
* speed_air, distance and duration have right skewed distribution
* Boeing has relatively higher distribution of distance and pitch than airbus
* Boeing has more speed_air variation. Mean and median are very close for both aircrafts
* Duration, No_pasg, Speed_ground and Height have similar distribution for both the aircrafts

__Initial analysis for identifying important factors that impact the response variable “landing distance”__

```{r}

# Compute the pairwise correlation between the landing distance and each factor X and storing it in Table1
cor1 <- cor(FAA_final$distance, FAA_final[,c(2,3,5,6)])

cor2 <- cor(subset(FAA_final,!is.na(FAA_final$speed_air))$distance,subset(FAA_final,!is.na(FAA_final$speed_air))$speed_air)


cor3 <- cor(subset(FAA_final,!is.na(FAA_final$duration))$distance,subset(FAA_final,!is.na(FAA_final$duration))$duration)

cor <- c(cor1,cor2,cor3)

direction <- c('negative','positive','positive','positive','positive','negative')

size_of_cor <- abs(cor)

names_variables <- colnames(FAA_final)[c(2,3,5,6,4,8)]

Table1 <- data.frame(names_variables,size_of_cor,direction)

Table1 <- Table1[
  order( Table1[,2],decreasing = TRUE ),
]

Table1

# Creating XY plots

plot(FAA_final$distance,FAA_final$duration)
plot(FAA_final$distance,FAA_final$no_pasg)
plot(FAA_final$distance,FAA_final$speed_ground)
plot(FAA_final$distance,FAA_final$speed_air)
plot(FAA_final$distance,FAA_final$height)
plot(FAA_final$distance,FAA_final$pitch)




```

On the basis of plots we observe some relationship between-

* Distance and Speed_ground
* Distance and Speed_air

The relationship is consistent with computed high correlations for these variables. Other variables do not show any linear relationship which is consistent with their low size of correlations.

```{r}
# Encoding aircraft (airbus is coded as 1 and boeing as 0)
for(i in 1:831)
{
if(FAA_final$aircraft[i] == 'airbus') 
{
  FAA_final$aircraft[i] = 1
}

if(FAA_final$aircraft[i] == 'boeing') 
{
  FAA_final$aircraft[i] = 0
}

}

FAA_final$aircraft <- as.numeric(FAA_final$aircraft)
cor_aircraft <- cor(FAA_final$aircraft,FAA_final$distance)

plot(FAA_final$aircraft,FAA_final$distance)





```

We can conclude from negative correlation and the plot that boeing has relatively higher distribution of distance than airbus.

__Regression using a single factor each time__

```{r}
lm_1 <- lm(distance ~ aircraft, FAA_final)
summary(lm_1)
d_1 <- 'negative'
p_1 <- summary(lm_1)$coefficients[,4][2]

lm_2 <- lm(distance ~ no_pasg, FAA_final)
summary(lm_2)
d_2 <- 'negative'
p_2 <- summary(lm_2)$coefficients[,4][2]

lm_3 <- lm(distance ~ speed_ground, FAA_final)
summary(lm_3)
d_3 <- 'positive'
p_3 <- summary(lm_3)$coefficients[,4][2]

lm_4 <- lm(distance ~ speed_air, FAA_final)
summary(lm_4)
d_4 <- 'positive'
p_4 <- summary(lm_4)$coefficients[,4][2]

lm_5 <- lm(distance ~ height, FAA_final)
summary(lm_5)
d_5 <- 'positive'
p_5 <- summary(lm_5)$coefficients[,4][2]

lm_6 <- lm(distance ~ pitch, FAA_final)
summary(lm_6)
d_6 <- 'positive'
p_6 <- summary(lm_6)$coefficients[,4][2]

lm_7 <- lm(distance ~ duration, FAA_final)
summary(lm_7)
d_7 <- 'negative'
p_7 <- summary(lm_7)$coefficients[,4][2]

p_value <- c(p_1,p_2,p_3,p_4,p_5,p_6,p_7)
dir <- c(d_1,d_2,d_3,d_4,d_5,d_6,d_7)
names <- colnames(FAA_final[c(1:6,8)])

Table2 <- data.frame(names,p_value,dir)
Table2 <- Table2[
  order( Table2[,2] ),
]
Table2

# Standardizing each X variable

m1 <- mean(FAA_final$aircraft)
s1 <- sd(FAA_final$aircraft)

m2 <- mean(FAA_final$no_pasg)
s2 <- sd(FAA_final$no_pasg)

m3 <- mean(FAA_final$speed_ground)
s3 <- sd(FAA_final$speed_ground)

m4 <- mean(FAA_final$speed_air, na.rm = TRUE)
s4 <- sd(FAA_final$speed_air, na.rm = TRUE)

m5 <- mean(FAA_final$height)
s5 <- sd(FAA_final$height)

m6 <- mean(FAA_final$pitch)
s6 <- sd(FAA_final$pitch)

m7 <- mean(FAA_final$duration, na.rm = TRUE)
s7 <- sd(FAA_final$duration, na.rm = TRUE)

FAA_final <- FAA_final %>% mutate(aircraft_s = (aircraft - m1)/s1)
FAA_final <- FAA_final %>% mutate(no_pasg_s = (no_pasg - m2)/s2)
FAA_final <- FAA_final %>% mutate(speed_ground_s = (speed_ground - m3)/s3)
FAA_final <- FAA_final %>% mutate(speed_air_s = (speed_air - m4)/s4)
FAA_final <- FAA_final %>% mutate(height_s = (height - m5)/s5)
FAA_final <- FAA_final %>% mutate(pitch_s = (pitch - m6)/s6)
FAA_final <- FAA_final %>% mutate(duration_s = (duration - m7)/s7)


lmm_1 <- lm(distance ~ aircraft_s, FAA_final)
summary(lmm_1)
dd_1 <- 'negative'
p_1 <- summary(lmm_1)$coefficients[,1][2]


lmm_2 <- lm(distance ~ no_pasg, FAA_final)
summary(lmm_2)
dd_2 <- 'negative'
p_2 <- summary(lmm_2)$coefficients[,1][2]

lmm_3 <- lm(distance ~ speed_ground, FAA_final)
summary(lmm_3)
dd_3 <- 'positive'
p_3 <- summary(lmm_3)$coefficients[,1][2]

lmm_4 <- lm(distance ~ speed_air, FAA_final)
summary(lmm_4)
dd_4 <- 'positive'
p_4 <- summary(lmm_4)$coefficients[,1][2]

lmm_5 <- lm(distance ~ height, FAA_final)
summary(lmm_5)
dd_5 <- 'positive'
p_5 <- summary(lmm_5)$coefficients[,1][2]

lmm_6 <- lm(distance ~ pitch, FAA_final)
summary(lmm_6)
dd_6 <- 'positive'
p_6 <- summary(lmm_6)$coefficients[,1][2]

lmm_7 <- lm(distance ~ duration, FAA_final)
summary(lmm_7)
dd_7 <- 'negative'
p_7 <- summary(lmm_7)$coefficients[,1][2]

size_coff <- c(p_1,p_2,p_3,p_4,p_5,p_6,p_7)
dir <- c(dd_1,dd_2,dd_3,dd_4,dd_5,dd_6,dd_7)
names <- colnames(FAA_final[c(9:15)])

Table3 <- data.frame(names,abs(size_coff),dir)
Table3 <- Table3[
  order( Table3[,2],decreasing = TRUE ),
]
Table3


```

__Relative order of importance on the basis of Table1,2 and 3__

Table1 and Table2 are consistent. However we observe some differences in the order in Table3.

```{r}
# Creating Table0 (Ranking based on all the factors)

names <- c('speed_ground','speed_air','aircraft','pitch','height','no_pasg','duration')
importance <- c(1:7)
Table0 <- data.frame(names,importance)
```

__Check collinearity__

```{r}
l1 <- lm(distance~speed_ground, FAA_final)
summary(l1)

l2 <- lm(distance~speed_air, FAA_final)
summary(l2)

l3 <- lm(distance~speed_ground + speed_air, FAA_final)
summary(l3)

```
 We observe sign change in speed_ground when we fit it along with speed_air in comparison when it is fitted alone.
 
```{r}
# Cor between speed_air and speed_ground

cor(subset(FAA_final,!is.na(FAA_final$speed_air))$speed_air,subset(FAA_final,!is.na(FAA_final$speed_air))$speed_ground)
```
 
 The correlation is very high. I would only keep speed_ground in the model as it does not have NAs. On the other hand speed_air has a lot of NAs.
 
 __Variable selection based on our ranking in Table 0__
 
```{r}
L1 <- lm(distance~speed_ground, FAA_final)
summary(L1)
r1 <- summary(L1)$r.squared

L2 <- lm(distance~speed_ground+aircraft, FAA_final)
summary(L2)
r2 <- summary(L2)$r.squared

L3 <- lm(distance~speed_ground+aircraft+pitch, FAA_final)
summary(L3)
r3 <- summary(L3)$r.squared

L4 <- lm(distance~speed_ground+aircraft+pitch+height, FAA_final)
summary(L4)
r4 <- summary(L4)$r.squared

L5 <- lm(distance~speed_ground+aircraft+pitch+height+no_pasg, FAA_final)
summary(L5)
r5 <- summary(L5)$r.squared

L6 <- lm(distance~speed_ground+aircraft+pitch+height+no_pasg+duration, FAA_final)
summary(L6)
r6 <- summary(L6)$r.squared

num <- c(1:6)

plot(c(r1,r2,r3,r4,r5,r6),num)

```
 
R squared increases as the number of variables increase

```{r}
adj.r1 <- summary(L1)$adj.r.squared
adj.r2 <- summary(L2)$adj.r.squared
adj.r3 <- summary(L3)$adj.r.squared
adj.r4 <- summary(L4)$adj.r.squared
adj.r5 <- summary(L5)$adj.r.squared
adj.r6 <- summary(L6)$adj.r.squared

c(adj.r1,adj.r2,adj.r3,adj.r4,adj.r5,adj.r6)
plot(c(adj.r1,adj.r2,adj.r3,adj.r4,adj.r5,adj.r6),num)

```

Adj R Squared also increases but the increase is less per addition of new variable

```{r}
# AIC

AIC1 <- AIC(L1)
AIC2 <- AIC(L2)
AIC3 <- AIC(L3)
AIC4 <- AIC(L4)
AIC5 <- AIC(L5)
AIC6 <- AIC(L6)

c(AIC1,AIC2,AIC3,AIC4,AIC5,AIC6)

plot(c(AIC1,AIC2,AIC3,AIC4,AIC5,AIC6),num)

```

On the basis of Adj R Squared and AIC I would select model L4. That is the model with 4 predictors - (speed_ground,aircraft,pitch and height)

__Variable selection based on automate algorithm__

```{r}
Model1_LM <- lm(distance ~ 1, data = FAA_final[,c(1:3,5:7)])
fit1_LM <- stepAIC(Model1_LM, direction = 'forward')

fit1 <- lm(distance ~ ., FAA_final[,c(1:3,5:7)])
fit2 <- lm(distance ~ 1, FAA_final[,c(1:3,5:7)])
stepAIC(fit1,direction="backward")
stepAIC(fit2,direction="forward",scope=list(upper=fit1,lower=fit2))
stepAIC(fit2,direction="both",scope=list(upper=fit1,lower=fit2))

```

On the basis of stepAIC forward variable selection I would select model with 2 predictors - (speed_ground and aircraft)
